---
title: "The little package that could"
subtitle: 'Taking visualizations to the next level with the scales package'
author: "Dana Seidel (@dpseidel)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [rladies, default-fonts, styles.css]
    nature:
      titleSlideClass: [center, middle, inverse]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---
class: middle

# What are scales really?? What is scaling? `r emo::ji("chart_with_upwards_trend")`
- converting data values to perceptual properties. 
- includes the inverse process: making guides (legends and axes) that can be used to read the graph 
Scaling and guides are often some of the most difficult parts of building any visualization.

---
class: center, middle

# Scales `r emo::ji("package")`
The scales package provides the internal scaling infrastructure to
[ggplot2](github.com/tidyverse/ggplot2) and exports standalone, system-agnostic,
functions. 

<strong> Use scales to customize 
the transformations, breaks, guides and palettes in your visualizations.
</strong>


---
class: center, middle

```{r setup, include=FALSE}
library(scales)
library(ggplot2)
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, dpi = 180, fig.height = 2, fig.width = 4)
```

# Highlighting some changes in scales 1.1.0

- naming conventions: 
- 
- 

---
class: center, middle, inverse
# Palettes

---
class: center, middle
# Colour palettes `r emo::ji("art")`
scales provides a number of color pallete functions that, given a range of values 
or the number of colours your want, will return a range of colors by hex code. 
```{r, palettes}
# pull a list of colours from any palette
viridis_pal()(4)
brewer_pal(type = "div", direction = -1)(4)
div_gradient_pal()(seq(0, 1, length.out = 4))
```

---
class: center, middle
# Color palettes (cont.)
```{r, fig.width = 2 }
# show_col is a quick way to view palette output
show_col(viridis_pal()(4))
```

---
class: center, middle
# Use scales palettes with baseR
These functions are primarily used under the hood in ggplot2, but can be combined 
with any plotting system. For example, use them in combination with `grDevices::palette()`,
provided with base R, to affect your base plots...

---
class: center, middle
# BaseR example
```{r, fig.height = 3}
palette(viridis_pal()(4))
plot(Sepal.Length ~ Sepal.Width, data = iris, col = Species, pch = 20)
```


---
class: center, middle
## Non-color palettes
Often you want to be able to scale elements other than color. e.g. size, alpha, shape... 
Of course, scales handles those too!
```{r}
your_data <- runif(13, 1, 20)
area_pal(range = c(1, 20))(your_data)
shape_pal()(6)
```

---
class: center, middle

# See these in action in ggplot2
```{r, eval=FALSE}
# color examples...
scale_fill_brewer()
scale_color_grey()
scale_color_viridis_c()
# shape examples
scale_shape()
scale_shape_ordinal()
# implement them yourself with...
scale_color_manual()
scale_shape_manual()
scale_size_manual()
# using available scales functions!
```

---
class: center, middle, inverse

# Guides & breaks

---
class: center, middle

# Making Guides easy: scales' formatters
The scales package also provides useful helper 
functions for formatting numeric data for all types of labels 
As of 1.0.0, most of scales formatters are just variations on the generic `number()`
and `number_format()` functions. 


---
class: center, middle

# The number formatter
+ The generic formatter offers consistency in behaviour and arguments across formatters
and adds additional functionality for international customization.
+ With it's adoption, existing formatters gained new, consistent arguments for 
`scale`, `accuracy`, `trim`, `big.mark`, `decimal.mark`, `prefix`, `suffix` etc. 

---
class: center, middle, inverse


# The number formatter, in action
By default, `number()` will take any numeric vector, round them to nearest whole
number, add spaces between every 3 digits and return a character vector useful for
feeding to a labels argument in ggplot2. 
```{r}
number(c(12.3, 4, 12345.789, 0.0002))
```

---
class: center, middle, inverse


# Changing defaults
You can easily specify a different rounding behavior, or change the `big_mark` 
or `decimal_mark` for international 
styling. Even add a `prefix` or a `suffix` or `scale` your numbers on the fly. 
```{r}
number(c(12.3, 4, 12345.789, 0.0002),
  big.mark = ".",
  decimal.mark = ",",
  accuracy = .01
)
```


---
class: center, middle, inverse

# Other Formatters:

`comma_format()` `comma()` `percent_format()` `percent()` `unit_format()`
`date_format()` `time_format()` : Formatted dates and times.
`dollar_format()` `dollar()` : Currency formatters, round to nearest cent and display dollar sign.
`ordinal_format()` `ordinal()` `ordinal_english()` `ordinal_french()` `ordinal_spanish()`: add ordinal suffixes (-st, -nd, -rd, -th) to numbers.
`pvalue_format()` `pvalue()` : p-values formatter
`scientific_format()` `scientific()` : Scientific formatter

---
class: center, middle, inverse


## Examples 
```{r, formatters}
# percent() function takes a numeric and does your division and labelling for you
percent(c(0.1, 1 / 3, 0.56))
# comma() adds commas into large numbers for easier readability
comma(10e6)
# dollar() adds currency symbols
dollar(c(100, 125, 3000))
# unit_format() adds unique units
# the scale argument allows for simple conversion on the fly
unit_format(unit = "ha", scale = 1e-4)(c(10e6, 10e4, 8e3))
```


---
class: center, middle, inverse

## Why the *_format() functions?
Where `number()` returns a character vector, `number_format()` and like functions
returns a fuctions that can be applied repeatedly or fed to a `labels` argument in a 
ggplot2 scale function. 
```{r}
# percent formatting in the French style
french_percent <- percent_format(decimal.mark = ",", suffix = " %")
french_percent(runif(10))
# currency formatting Euros (and simple conversion!)
usd_to_euro <- dollar_format(prefix = "", suffix = "\u20ac", scale = .86)
usd_to_euro(100)
```


---
class: center, middle, inverse

## Applied in ggplot scales
```{r}
dsamp <- dplyr::sample_n(diamonds, 1000)
ggplot(dsamp, aes(x = carat, y = price, colour = clarity)) +
  geom_point() + scale_y_continuous(labels = usd_to_euro)
```

---
class: center, middle, inverse


## Breaks
<font size="4">
`scales::extended_breaks()` sets most breaks by default in ggplot2
`pretty_breaks()` is an alternative break calculation 
Many of the formatter and transformation functions have matching break functions, eg:
- `log_breaks()` is used to set breaks for log transformed axes with `log_trans()`.
- `date_breaks()` is used to set nice breaks for date and date/time axes.
</font>

---
class: center, middle


# Bounds & transformations

---
class: center, middle, inverse


## Rescaling data
scales provides a handful of functions for rescaling data to fit new ranges.
```{r rescale}
# the rescale functions can rescale continuous vectors to new min, mid, or max values
x <- runif(5, 0, 1)
x
rescale(x, to = c(0, 50))
rescale_mid(x, mid = .25)
rescale_max(x, to = c(0, 50))
```

---
class: center, middle, inverse

## Squish, Discard, Censor
```{r squish}
# squish() will squish your values into a specified range, respecting NAs
squish(c(-1, 0.5, 1, 2, NA), range = c(0, 1))
# discard will drop data outside a range, respecting NAs
scales::discard(c(-1, 0.5, 1, 2, NA), range = c(0, 1))
# censor will return NAs for values outside a range
censor(c(-1, 0.5, 1, 2, NA), range = c(0, 1))
```

---
class: center, middle, inverse

## Applied to ggplot2
<font size= "4">Squish can be really useful for setting the `oob` argument for a colour scale with reduced limits.</font>
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Sepal.Length)) +
  geom_point() + scale_color_continuous(limit = c(6, 8), oob = scales::squish)
```

---
class: center, middle, inverse

## Transformations
scales provides a number of common transformation functions (`*_trans()`) which 
specify functions to preform data transformations, format labels, and set correct breaks.
For example:
`log_trans()`, `sqrt_trans()`, `reverse_trans()` power the `scale_*_log10()`,
`scale_*_sqrt()`, `scale_*_reverse()` functions in ggplot2. 

---
class: center, middle, inverse

## Additional Transformations
- `asn_trans()`: Arc-sin square root transformation.
- `atanh_trans()`: Arc-tangent transformation.
- `boxcox_trans()` `modulus_trans()` : Box-Cox & modulus transformations.
- `date_trans()` `time_trans()` `hms_trans()` : transformations for date, datetime, and hms classes
- `exp_trans()` : Exponential transformation (inverse of log transformation).
- `pseudolog_trans()` : Pseudo-log transformation
- `probabilty_trans()`: Probability transformation
and more... 


---
class: center, middle, inverse

## Building your own transformations
scales also gives users the ability to define and apply their own custom 
transformation functions for repeated use.
```{r transforms}
# use trans_new to build a new transformation
dollar_log <- trans_new(
  name = "dollar_log",
  trans = log_trans(base = 10)$trans, # extract a single element from another trans
  inverse = function(x) 10^(x), # or write your own custom functions
  breaks = log_breaks(),
  format = dollar_format()
  )
```

---
class: center, middle, inverse

## Applied in ggplot2
```{r}
# apply our new transformation!
ggplot(dsamp, aes(x = carat, y = price, colour = clarity)) +
  geom_point() + scale_y_continuous(trans = dollar_log)
```

---
class: center, middle, inverse

## Questions?

Slides: [danaseidel.com/rstudioconf2020](https://danaseidel.com/rstudioconf2020)

Code: [github.com/dpseidel/rstudioconf2020](https://github.com/dpseidel/rstudioconf2020)

@dpseidel on [Twitter](https://twitter.com/dpseidel) and [GitHub](https://github.com/dpseidel)